Accesible, realistic genome simulation with selection using \stdpopsim

### 2025-03-12 analysis2_manuscript.tex
one more DFE method edit
more methods

    \subsection*{Whole genome simulations}
    In this study we performed simulations of the human and vaquita porpoise genomes
    using the \stdpopsim{} framework.
    All simulations were executed using the \slim engine with a burn-in of $2N$ generations
    (where $N$ is the initial population size), and a scaling factor of 2 (to reduce running time).
    Each simulation was implemented using a relatively short python script, similar to the one shown in Figure 1B
    (see \textbf{Data availablity} below).
    \igcomment{We should add a link to the analysis2 repo in data/code availablity section}
    %https://github.com/popsim-consortium/analysis2/tree/main
    % DRS: Maybe we should say something about how easy it was to do these simulations, and maybe even show
    % some example code somewhere to prove our point? Could help emphasize the utility of this resource.
    % IG: added a note about this above.

    \subsubsection*{Simulations of complete human genomes}
    Simulation of human genomes used a genetic map from the HapMap Project
    \citep{international2007second} (\stdpopsim label \texttt{HapMapII\_GRCh38}) and
    the out of Africa (OOA) demographic model with archaic admixture from
    \cite{ragsdale2019models} (\stdpopsim label \texttt{OutOfAfricaArchaicAdmixture\_5R19}).
    Simulations with selection additionally applied a DFE model with gamma-distributed selection coefficients inferred by
    \cite{kim2017inference} (\stdpopsim label \texttt{Gamma\_K17}) to exons annotated by the HAVANA group release 104
    (\cite{ensembl2018}; \stdpopsim label \texttt{ensembl\_havana\_104\_exons}).
    Each simulation generated 100 diploid genomes from each of the three extant human populations (YRI, CEU, and CHB),
    with each genome consisting of 22 autosomes.
    Using these settings, we generated three replicate datasets without selection
    and three replicate datasets with selection.
    Three additional datasets, used in Figure \ref{fig:homsap-dfe.constant},
    were simulated with the same model of selection and a simple demographic model with one population with a constant size of $10,000$.
    Each of these simulations generated 100 diploid genomes.

    \subsubsection*{Simulations of complete vaquita porpoise genomes}
    Simulation of vaquita porpoise genomes used 
    the two-epoch demographic model from \cite{robinson2022critically} (\stdpopsim label \texttt{Vaquita2Epoch\_1R22}).
    Since no genetic map is currently available for this species, we assumed a constant recombination rate of  $10^{-8}$
    across all chromosomes \citep{morin2021}.
    Simulations with selection additionally applied a DFE model with gamma-distributed selection coefficients inferred by
    \cite{robinson2022critically} (\stdpopsim label \texttt{Gamma\_R22}) to exons annotated from the vaquita genome assembly
    (\cite{morin2021}; \stdpopsim label \texttt{Phocoena\_sinus.mPhoSin1.pri.110\_exons}).
    This DFE model implements a relationship between selection and dominance as follows:
    very deleterious mutations (with $s<-0.1$) were set to be recessive ($h=0$),
    mutations with $s\in [-0.1,-0.01)$ were set to have a dominance coefficient of $h=0.01$,
    mutations with $s\in [-0.01,-0.001)$ were set to have a dominance coefficient of $h=0.1$,
    and very mild mutations (with $s\in [-0.001,0)$) were set to have a nearly additive effect ($h=0.4$).
    This approach is similar to what was used in the simulations cunducted by \textcite{robinson2022critically}
    and is easily implemented in \stdpopsim's \texttt{DFE} class.
    Each simulation generate 100 diploid genomes,
    with each genome consisting of 21 autosomes.
    Using these settings, we generated three replicate datasets without selection
    and three replicate datasets with selection.
    Three additional datasets, used in Figure \ref{fig:vaquita-dfe.constant},
    were simulated with the same model of selection and a simple demographic model with one population with a constant size of $3,500$.
    \igcomment{If we end up using Fig. S3, then we should add a line describing it's simulation here.}

    \subsection*{Inference of $N_e(t)$}
    %We assessed the accuracy of demographic inferences from
    %\msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.
    Inference of $N_e(t)$ was conducted on simulated datasets after masking non-recombining portion
    of the genome (to avoid bias caused by extreme recombination rate variation)
    and exons (to avoid the direct effect of natural selection).
    Thus, inference from datasets simulated with selection was not influenced by direct selection
    acting on exons, but it was influenced by selection acting on linked sites (background selection).    
    %IG: I commented because it doesn't seem relevant for methods
    %Comparisons between inferences made with and without selection were 
    %of particular interest.
    For \msmc \citep{Schiffels2020}, we used a random sample of $N=6$ simulated genomes with 20
    iterations of the EM algorithm.
    For \gone \citep{santiago2020recent}, we set \texttt{max\_snps} to 500,000,
    the number of generations to 2,000, and the number of bins to 400, all representing default settings.
    For \stairway \citep{liu2020stairway} and \smcpp \citep{terhorst2017robust} we used the default settings.
    \igcomment{Nothing more to say about \stairway plot and \smcpp?}

    \subsection*{DFE inference}
    \igcomment{I commented all the text that was here because it wasn't relevant to the methods. We might want to add it to the discussion.
    Kirk also noted that a lot of methods-level details are missing:
    how big was the coding sequence length?
    How was demographic inference done for the different models via the different programs?,
    etc.
    Can someone write this up?}
    % ADK: I've added a section below that describes the DFE inference methods used in this study.

@@

    % ADK: I've added a section below that describes the DFE inference methods used in this study.
    For comparison among methods, estimates were rescaled. 
    As described above, each DFE-inference method scales selection coefficients slightly differently.
    We thus rescaled each program's output to obtain a selection coefficient ($s$) consistent with the one used by \slim.
    The output of \grapes  was scaled by a factor of  $\mu / \theta$,
    and the output of \polydfe was saled by $2 \mu / \theta$, 
    where $\mu$ is the assumed mutation rate and 
    $\theta$ is the population-scaled mutation rate estimated by each method ($\theta = 4N_{e}\mu$).
    For \dadicli, we report $\alpha * \nu / (\theta / (4 \mu L p))$, where
    $\alpha$ and $\nu$ are the inferred shape and rate parameters of the \dadicli's inferred gamma distribution,
    $L$ is the total number of sites available for the analysis,
    $p$ is the proportion of sites under selection.

@@

    \subsection*{Selective sweeps detection}
    %\drscomment{Maybe it makes sense to move this up to right after the sweep simulation section. Should make it easier for readers to follow.}
    %IG: I moved that section down here to bring them together

    Each of the three sweep-detection methods is based on a test statistic.
    Reduced nucleotide diversity is based on nucleotide diversity  ($\pi$) 
    computed directly from the underlying tree sequences using \tskit \citep{ralph2020efficiently}
    over 10 equally sized and non-overlapping subwindows.
    \sweepfinder \citep{degiorgio2016sweepfinder2} is based a composite likelihood ratio (CLR) 
    computed at 21 uniformly distributed positions along the simulated region.
    \diploshic \citep{kern2018diplos} uses supervised learning to train a sweep classifier.
    Its training phase used \texttt{discoal} \citep{kern2016discoal} to simulate 20 haploid sequences of length 1.21Mbp
    with a constant population size of $N_e=10,000$, a mutation rate of $2\times 10^{-8}$, and a recombination rate of $2\times 10^{-8}$.
    For \texttt{discoal} simulations that included selective sweeps,
    selection coefficients were drawn from a uniform distribution between 0.001 and 0.10,
    and the time of the sweep was chosen from a uniform distribution between 0 and $0.01N_e$ generations in the past.
    For soft sweeps, the frequency at which selection was introduced was drawn from a uniform distribution between $0.0001$ and $0.1$.
    We note that since the training set of \diploshic was generated under a simple model with
    constant mutation and recombination rates and constant effective population size,
    we expect it to perform suboptimally on the simulated segments of human chromosome 1,
    which were generated using more relaistic simulations (see \textbf{Simulations with selective sweeps} above).

    The three sweep-detection methods were associated with thresholds for their respective test statistics
    ($\pi$, \sweepfinder's CLR, or \diploshic's probability),
    which were determined using an empirical distribution representing a null model without sweeps.
    We considered two null models: a neutral evolution model (neutral),
    and a model with background selection from deleterious mutations in exons (BGS).
    Each null model was represented by a different set of $100,000$ simulations generated according to the appropriate model
    (see \textbf{Simulations with selective sweeps} above).
    Thus, for each test statistic we determined two thresholds corresponding to these two different null models (neutral and BGS)
    using the extreme $5\%$ quantile in each empirical distribution (and excluding regions with extremely low recombination rate).
    \igcomment{Can we say what the rec rate cutoff was and how many regions were excluded here?
    Shouldn't we have excluded these regions from the simulations altogether?}.
    Then, power (or true positive rate) to detect a sweep by a given method was determined for
    100 evenly-spaced positions on chromosome 1 as follows.
    The sweep-detection method was applied to the 1,000 
    simulated datasets generated with a hard sweep (neutral and BGS) for that position,
    and the power was set to be the proportion of simulations (out of 1,000)
    in which the test statistic exceeded the threshold determined by each of the two null models (neutral and BGS).

>>>>

    \subsection*{Whole genome simulations}
    In this study we performed simulations of the human and vaquita porpoise genomes
    using the \stdpopsim{} framework. We ran replicate datasets for each genome/parameter combination,
    in which chromosomes were simulated separately from one another.
    All simulations were executed using the \slim engine with a burn-in of $10N$ generations
    (where $N$ is the initial population size), and a scaling factor of 1 (no scaling).
    All simulations were implemented using as part of a snakemake workflow,
    where the code utilized is similar to the one shown in Figure 1B.
    The complete simulation workflow is available at \url{https://github.com/popsim-consortium/analysis2/blob/main/workflows/simulation.snake}
    (see \textbf{Data availablity} below).

    \subsubsection*{Simulations of complete human genomes}
    Simulation of human genomes used a genetic map from the HapMap Project
    \citep{international2007second} (\stdpopsim label \texttt{HapMapII\_GRCh38}) and
    the out of Africa (OOA) demographic model with archaic admixture from
    \cite{ragsdale2019models} (\stdpopsim label \texttt{OutOfAfricaArchaicAdmixture\_5R19}).
    Simulations with selection additionally applied a DFE model with gamma-distributed selection coefficients inferred by
    \cite{kim2017inference} (\stdpopsim label \texttt{Gamma\_K17}) to exons annotated by the HAVANA group release 104
    (\cite{ensembl2018}; \stdpopsim label \texttt{ensembl\_havana\_104\_exons}).
    Each simulation generated 100 diploid genomes from each of the three extant human populations (YRI, CEU, and CHB),
    with each genome consisting of 22 autosomes.
    Using these settings, we generated three replicate datasets without selection
    and three replicate datasets with selection.
    Three additional datasets, used in Figure \ref{fig:homsap-dfe.constant},
    were simulated with the same model of selection and a simple demographic model with one population with a constant size of $10,000$.
    Each of these simulations generated 100 diploid genomes.
    The configuration file used in combination with our snakemake workflow
    for these simulations can be found at \url{https://github.com/popsim-consortium/analysis2/blob/main/workflows/config/snakemake/production_config_HomSap.yml}.

    \subsubsection*{Simulations of complete vaquita porpoise genomes}
    Simulation of vaquita porpoise genomes used 
    the two-epoch demographic model from \cite{robinson2022critically} (\stdpopsim label \texttt{Vaquita2Epoch\_1R22}).
    Since no genetic map is currently available for this species, we assumed a constant recombination rate of  $10^{-8}$
    across all chromosomes \citep{morin2021}.
    Simulations with selection additionally applied a DFE model with gamma-distributed selection coefficients inferred by
    \cite{robinson2022critically} (\stdpopsim label \texttt{Gamma\_R22}) to exons annotated from the vaquita genome assembly
    (\cite{morin2021}; \stdpopsim label \texttt{Phocoena\_sinus.mPhoSin1.pri.110\_exons}).
    This DFE model implements a relationship between selection and dominance as follows:
    very deleterious mutations (with $s<-0.1$) were set to be recessive ($h=0$),
    mutations with $s\in [-0.1,-0.01)$ were set to have a dominance coefficient of $h=0.01$,
    mutations with $s\in [-0.01,-0.001)$ were set to have a dominance coefficient of $h=0.1$,
    and very mild mutations (with $s\in [-0.001,0)$) were set to have a nearly additive effect ($h=0.4$).
    This approach is similar to what was used in the simulations cunducted by \textcite{robinson2022critically}
    and is easily implemented in \stdpopsim's \texttt{DFE} class.
    Each simulation generate 100 diploid genomes,
    with each genome consisting of 21 autosomes.
    Using these settings, we generated three replicate datasets without selection
    and three replicate datasets with selection.
    Three additional datasets, used in Figure \ref{fig:vaquita-dfe.constant},
    were simulated with the same model of selection and a simple demographic model with one population with a constant size of $3,500$.
    \igcomment{If we end up using Fig. S3, then we should add a line describing it's simulation here.}
    The configuration file used in combination with our snakemake workflow
    for these simulations can be found at 
    \url{https://github.com/popsim-consortium/analysis2/blob/main/workflows/config/snakemake/production_config_PhoSin.yml}.

    \subsection*{Inference of $N_e(t)$}
    Inference of $N_e(t)$ was conducted on simulated datasets after masking non-recombining portions
    of the genome (to avoid bias caused by extreme recombination rate variation)
    and exons (to avoid the direct effect of natural selection).
    Thus, inference from datasets simulated with selection was not influenced by direct selection
    acting on exons, but it was influenced by selection acting on linked sites (background selection).

    We assessed the accuracy of demographic inferences from three software packages:
    \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent},
     and \smcpp \citep{terhorst2017robust}.
    For \msmc \citep{Schiffels2020}, we used a random sample of $N=6$ simulated genomes with 20
    iterations of the EM algorithm.
    For \gone \citep{santiago2020recent}, we set \texttt{max\_snps} to 500,000,
    the number of generations to 2,000, and the number of bins to 400, all representing default settings.
    For \stairway \citep{liu2020stairway} we did one run of theta estimation for each replicate dataset,
    and used a `dimFactor' of 5000, both of which represent default settings.
    For \smcpp \citep{terhorst2017robust} we used the unfolded SFS (`--unfold'), 
    along with the true mutation rate under which the simulations were generated. 
    The complete snakemake workflow for demographic inference can be found at
    \url{https://github.com/popsim-consortium/analysis2/blob/main/workflows/n_t.snake}.

    \subsection*{DFE inference}
    %\igcomment{I commented all the text that was here because it wasn't relevant to the methods. We might want to add it to the discussion.
    %Kirk also noted that a lot of methods-level details are missing:
    %how big was the coding sequence length?
    %How was demographic inference done for the different models via the different programs?,
    %etc.
    %Can someone write this up?
    % ADK: I've added a section below that describes the DFE inference methods used in this study.

@@

    % ADK: I've added a section below that describes the DFE inference methods used in this study.
    For comparison among methods, estimates were rescaled. 
    As described above, each DFE-inference method scales selection coefficients slightly differently.
    We thus rescaled each program's output to obtain a selection coefficient ($s$) consistent with the one used by \slim.
    The output of \grapes  was scaled by a factor of  $\mu / \theta$,
    and the output of \polydfe was saled by $2 \mu / \theta$, 
    where $\mu$ is the assumed mutation rate and 
    $\theta$ is the population-scaled mutation rate estimated by each method ($\theta = 4N_{e}\mu$).
    For \dadicli, we report $\alpha * \nu / (\theta / (4 \mu L p))$, where
    $\alpha$ and $\nu$ are the inferred shape and rate parameters of the \dadicli's inferred gamma distribution,
    $L$ is the total number of sites available for the analysis,
    $p$ is the proportion of sites under selection. 
    Complete details for the DFE inference analysis can be found in the
    snakemake workflow at \url{https://github.com/popsim-consortium/analysis2/blob/main/workflows/dfe.snake}.

@@

    \subsection*{Selective sweeps detection}

    Each of the three sweep-detection methods is based on a test statistic.
    Reduced nucleotide diversity is based on nucleotide diversity  ($\pi$) 
    computed directly from the underlying tree sequences using \tskit \citep{ralph2020efficiently}
    over 10 equally sized and non-overlapping subwindows.
    \sweepfinder \citep{degiorgio2016sweepfinder2} is based on a composite likelihood ratio (CLR) 
    computed at 21 uniformly distributed positions along the simulated region.
    \diploshic \citep{kern2018diplos} uses supervised learning to train a sweep classifier.
    Its training phase used \texttt{discoal} \citep{kern2016discoal} to simulate 20 haploid sequences of length 1.21Mbp
    with a constant population size of $N_e=10,000$, a mutation rate of $2\times 10^{-8}$, and a recombination rate of $2\times 10^{-8}$.
    For \texttt{discoal} simulations that included selective sweeps,
    selection coefficients were drawn from a uniform distribution between $0.001$ and $0.10$,
    and the time of the sweep was chosen from a uniform distribution between $0$ and $0.01N_e$ generations in the past.
    For soft sweeps, the frequency at which selection was introduced was drawn from a uniform distribution between $0.0001$ and $0.1$.
    We note that since the training set of \diploshic was generated under a simple model with
    constant mutation and recombination rates and constant effective population size,
    we expect it to perform suboptimally on the simulated segments of human chromosome 1,
    which were generated using more relaistic simulations (see \textbf{Simulations with selective sweeps} above).

    The three sweep-detection methods were associated with thresholds for their respective test statistics
    ($\pi$, \sweepfinder's CLR, or \diploshic's probability),
    which were determined using an empirical distribution representing a null model without sweeps.
    We considered two null models: a neutral evolution model (neutral),
    and a model with background selection from deleterious mutations in exons (BGS).
    Each null model was represented by a different set of $100,000$ simulations generated according to the appropriate model
    (see \textbf{Simulations with selective sweeps} above).
    Thus, for each test statistic we determined two thresholds corresponding to these two different null models (neutral and BGS)
    using the extreme $5\%$ quantile in each empirical distribution (and excluding regions with extremely low recombination rate).
    \igcomment{Can we say what the rec rate cutoff was and how many regions were excluded here?
    Shouldn't we have excluded these regions from the simulations altogether?}.
    Then, power (or true positive rate) to detect a sweep by a given method was determined for
    100 evenly-spaced positions on chromosome 1 as follows:
    each sweep-detection method was applied to the 1,000 
    simulated datasets generated with a hard sweep (neutral and BGS) for that position,
    and the power was set to be the proportion of simulations (out of 1,000)
    in which the test statistic exceeded the threshold determined by each of the two null models (neutral and BGS).
    We note that using an empirical distribution to determine the threshold for calling a selective
    sweep with \diploshic deviates from the original implementation of the method,
    however for consistency among methods we chose to take this approach.
    The complete snakemake workflow for sweep detection can be found at
    \url{https://github.com/popsim-consortium/analysis2/blob/main/workflows/sweep_simulate.snake}.
